---
title: "R Workshop"
author: "Aurélien Ginolhac"
date: "15 March 2016"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Dear all,

I would like to organize a workshop for the LSRU' people who want to learn / improve their _R_ skills.
Starting from scratch a _R_ course does not seem relevant neither effective.  
On the contrary, learning from concrete examples and focusing on modern packages should help.

If you are interested, you can continue reading. After the description, you can find a link to
a survey so I can adapt the content and time dedicated.

## Modern packages

What I mean is that _R_ base by itself is harsh and far from intuitive.
We want to avoid those kind of coding, even if it's working fine
```{r, eval=FALSE}
# from https://github.com/ANGSD/angsd/blob/master/R/jackKnife.R
Args<-function(l,args){
 if(! all(sapply(strsplit(l,"="),function(x)x[1])%in%names(args))){
  cat("Error -> ",l[!sapply(strsplit(l,"="),function(x)x[1])%in%names(args)]," is not a valid argument")
  q("no")
}
```


The libraries for the **main** components are:

+ **reading** data, get rid of `read.table()`
    - `readr` for all text files, much faster, much better
    - `readxl` for excel files. Fast, still some issues but overall great
+ **tidying** data. The main concept is simple:
    - Each column is a variable.
    - Each row is an observation.
    - `tidyr` does the job
+ **manipulating** data. Without entering into details, everything like merging, filtering, sorting, aggregating etc.
    - `dplyr`
+ **plotting** data. _R_ does an amazing job
    - `ggplot2`
+ **chaining** functions. Sick of parenthesis, we will move to the pipe operator `%>%` 

Depending on the interest, the 4 following aspects could also be covered:

+ **modeling**, where _R_ base actually is great.
    - `broom` for cleaning up models
+ **writing** nice reports
   - `Rmarkdown` and `knitr` (this very document)
+ **versioning**, using `git`. Not really specific to _R_ but nicely integrated into RStudio.   
+ **interactive** computation, Web framework.
    - `shiny` and related packages. See this example developed by [Eric Koncina](http://shiny-server.uni.lu/elisar-shiny/)
  
Of course, everything will be done using [Rstudio](https://www.rstudio.com/products/RStudio/#Desktop) that makes your life way easier.


## Concrete examples

```{r, include=FALSE}
library("ggplot2")
library("scales")
library("tidyr")
library("readr")
library("readxl")
library("dplyr", warn.conflicts = FALSE)
theme_set(theme_bw(14))
```


### merging list / tables 


A common question is _how do I merge 2 lists based on common id(s)?_

Let's consider the two following and distinct datasets, *a* and *b*

```{r}
library("dplyr")
a <- data.frame(id = 1:3,
                name = c("a", "a", "b"))
b <- data.frame(id = c(2, 3, 1),
                other = c("yes", "no", "yes"))
a
b
```

Obviously, the ids are not sorted the same way, but with R, it is very easy to merge them by id:

```{r}
inner_join(a, b)
```

<!-- the base version is `merge` which is ~ 50 times slower 
 Je ne pense pas que ce soit utile pour le pitch... Il y a déjà beaucoup de mots clés "pointus": Tout dépend du public visé...
-->

## Tidying

A dataset is considered tidy, once each observation is in its own row and each variable in its own column.

```{r, echo = FALSE}
gene_expression <- structure(c(5.85816859103934, 6.60428769475497, 5.4278181170489, 
3.64198318686737, 2.98308007124715, 6.60444574589028, 7.00805674362776, 
5.57135166957136, 5.14045415430309, 3.49663423366768, 5.94121052204627, 
7.25262696967957, 5.73597527709437, 5.13355247068666, 3.09661935730267, 
8.19815765591581, 6.41420670534176, 6.2578760673826, 4.97723524689115, 
1.90301643803504, 6.79101402711924, 7.03201225669769, 5.68340149938261, 
4.67483151462603, 3.78712626322116, 5.8756526105461, 7.89718752225727, 
5.58581882538314, 3.95586533543214, 4.29670995199073, 5.72651601739691, 
7.14934863678698, 5.34678828849904, 5.47530710160199, 5.79145434700723, 
5.19431908120944, 7.07334883453955, 5.56747268379296, 4.67120897018278, 
4.86218289261696), .Dim = c(5L, 8L), .Dimnames = list(c("10000_at", 
"10001_at", "10002_at", "100033423_at", "100033424_at"), c("GSM875933", 
"GSM875934", "GSM875935", "GSM875936", "GSM875937", "GSM875938", 
"GSM875939", "GSM875940"))) %>% as.data.frame()
```

In the following example, each sample is represented in its column.

```{r}
gene_expression
```

Samples ending with numbers 3, 4 or 5 are the control conditions while the remaining are the treated one. The above _wide_ format representation is not very handy to use. We can easily tidy it (i.e. transform it into the _long_ format):

```{r}
library("tidyr")
tidy_ge <- gather(gene_expression, sample, value) %>% # chain to tbl_df
  tbl_df()  # that format nicely the table printing, 10 instead of the full 40 rows
tidy_ge
```

Once our data is tidy, it is very easy to improve it and add additional useful informations. Let us specify which samples are controls and which are treated: 

```{r}
tidy_ge <- tidy_ge %>%
  mutate(condition = ifelse(grepl("[345]$", sample), "control", "treated"))
tidy_ge
```


## Summarizing

Let's use the previous example to compute the mean and standard deviation for each sample and condition:

```{r}
sum_ge <- tidy_ge %>%
  group_by(sample, condition) %>%
  summarise(mean_ge = mean(value),
            sd_ge = sd(value))
sum_ge
```


## Plotting

### Intervals

Since our previous summarised example data is **tidy**, we can use it directly (see that the legend is automatically added).

```{r, fig.height=5, fig.width=4}
library("ggplot2")
sum_ge %>%
  ggplot(aes(x = sample, y = mean_ge, colour = condition))+
  geom_pointrange(aes(ymin = mean_ge - sd_ge,
                      ymax = mean_ge + sd_ge))+
  coord_flip()
```

If you prefer the old companion barplot

```{r, fig.height=4, fig.width=6}
sum_ge %>%
  ggplot(aes(x = sample, y = mean_ge, fill = condition))+
  geom_bar(stat = "identity")+
  geom_errorbar(aes(ymin = mean_ge - sd_ge,
                      ymax = mean_ge + sd_ge), width = 0.5)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### Scatterplot and linear regression

```{r, fig.height=5, fig.width=6}
iris %>% # use built-in dataset
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, colour = Species))+ # declare the mapped variables
  geom_point()+ # add points
  geom_smooth(method = "lm") # add linear regression, automatically per Species
```


## Questionnaire

You can access the questionr [here](http://goo.gl/forms/qcU6Qvgmw3)
